Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-09 rnni_convexity.tex
proof of lemma 4 is problematic

For proving the other direction of the statement, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
We prove that $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$. The induction basis for $n=3$ taxa is trivial as the $\rnni$ graph is a triangle on three caterpillar trees.

For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree at distance $\frac{n(n-1)}{2}$ to $T$.
It follows that when applying $\csort$ onto $T,R$ the number of taxon swaps is maximum in each step, which implies that taxon $n+1$ is part of the cherry of $R$.
Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same $\nni$ moves as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.
With Lemma~\ref{lemma:distance_delete_taxon} we also know that it is $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = d(T,R) - (n-1)$, where the last equality follows from our observation above.

@@

\begin{theorem}
The set of caterpillar trees is convex.
\label{thm:caterpillar_convex}
\end{theorem}

>>>>

For proving the other direction of the statement, assume that $d_c(T,R) = \frac{(n-1)(n-2)}{2}$.
We prove that $d(T,R) = \frac{(n-1)(n-2)}{2}$ by induction on the number of taxa $n$.
The induction basis for $n=3$ taxa is trivial as the $\rnni$ graph is a triangle on three caterpillar trees in that case.

For the induction step we assume without loss of generality that $T$ is the caterpillar tree $[1,2,3,\ldots,n+1]$ and $R$ is a caterpillar tree at distance $\frac{n(n-1)}{2}$ to $T$.
\todo{We cannot assume that and the rest of the proof doesn't go through.}
It follows that when applying $\csort$ onto $T,R$ the number of taxon swaps is maximum in each step, which implies that taxon $n+1$ is part of the cherry of $R$.
Let us now consider the trees $T{\big|}_n$ to $R{\big|}_n$ that result from $T$ and $R$ by deleting taxon $n+1$.
$\csort$ computes a path from $T{\big|}_n$ to $R{\big|}_n$ that contains the same $\nni$ moves as $p$, except for the $n-1$ ones on $p$ that involve taxon $n+1$.
Hence, $d_c(T{\big|}_n, R{\big|}_n) = \frac{n(n-1)}{2} - (n-1)$ = $\frac{(n-1)(n-2)}{2}$.
So we can apply the induction hypothesis on $T{\big|}_n$ and $R{\big|}_n$ and know that $d(T{\big|}_n,R{\big|}_n) = \frac{(n-1)(n-2)}{2}$.
With Lemma~\ref{lemma:distance_delete_taxon} we also know that it is $d(T{\big|}_n, R{\big|}_n) \leq d(T,R) - |\rank_T(\parent_T(n+1)) - \rank_R(\parent_R(n+1))| = d(T,R) - (n-1)$, where the last equality follows from our observation above.

@@

\begin{theorem}
The set of caterpillar trees is convex in $\rnni$.
\label{thm:caterpillar_convex}
\end{theorem}
